Natural Selection Across Three Generations of Americans

### 2024-01-08 hrs-selection.qmd
TE update Jan 8

---
title: "Natural selection in the Health and Retirement Study"
author: 
  - David Hugh-Jones
  - Tobias Edwards
abstract: |
  We investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  partially support the economic theory of fertility as an explanation for
  scores which correlate negatively (positively) with education are
  selected for (against). Selection coefficients are larger among low-income
  and unmarried parents, but not among younger parents or those with less 
  education. We also estimate effect sizes corrected for noise in the 
  polygenic scores. 
  natural selection: among both white and black respondents,
date: "November 2023"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

```{r}
#| label: setup

# TE quick TODOs change reference to time being insignificant. Fix the footnote citations
options(digits = 3)
set.seed(271075)

@@

# TE edit PGI repository EUR only
pgirep <- readr::read_tsv("data/pgi-repository/HRS_PGIrepo_v1.0.txt",
                          show_col_types = FALSE)
names(pgirep)[1:2] <- c("hhid","pn")

@@

# rand family respondents file
# Done just once:
# rand_fam <- haven::read_dta(
#   "data/randhrsfam1992_2018v2_STATA/randhrsfamr1992_2018v2.dta")
# saveRDS(rand_fam, "data/randhrsfam1992_2018v2_STATA/rand_fam.rds")
rand_fam <- readRDS("data/randhrsfam1992_2018v2_STATA/rand_fam.rds")
# only selected columns to avoid creating suffixes when we join with
# rand.
# Age of oldest kid; grandchildren; num siblings who died
rand_fam <- rand_fam |> select(hhid, pn, h10ageokid, h10gkid, r10sbdied)

@@

# TE I'm replacing 0 weights with NA because it can create odd problems
rand$weight <- na_if(rand$weight, 0)

@@

# TE this is average number of kids that your kids have
rand$div_h10gkid <- rand$h10gkid / rand$raevbrn

# 83 people report having grandkids, but having had no children. Weird!
# This code subsets the data by ethnicity then counts all observations
# where grandkids is greater than 0 but number of kids is 0
gkids_unusable_white <- rand |> 
  filter(ethnicity == "white", h10gkid > 0, raevbrn == 0) |> 
  nrow()
gkids_unusable_black <- rand |> 
  filter(ethnicity == "black", h10gkid > 0, raevbrn == 0) |> 
  nrow()

# if grandkids but no kids div_gkid is infinite. Replacing with NA
rand$div_h10gkid[is.infinite(rand$div_h10gkid)] <- NA

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_weight = mean(weight, na.rm = TRUE),
    mean_child_weight = mean(child_weight, na.rm = TRUE),
    mean_raevbrn = mean(raevbrn*weight/mean_weight, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn,
    mean_h10gkid = mean(h10gkid*weight/mean_weight, na.rm = TRUE),
    rlrs_h10gkid = h10gkid/mean_h10gkid,
    mean_div_h10gkid = mean(div_h10gkid*child_weight/mean_child_weight, 
                            na.rm = TRUE), 
    rlrs_div_h10gkid = div_h10gkid/mean_div_h10gkid
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    r10hearte = as.numeric(r10hearte),
    r10diabe = as.numeric(r10diabe),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    born = chop(rabyear, 1942, labels = c("pre-1942", "post-1942")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )

# TE I've plotted the mean values by year. They look reasonably stable. Beauchamp 2016 took a three year average. I don't think this is necessary

@@

rand <- rand |> 
  select(ethnicity, all_of(pgs), all_of(old_pgs), 
         all_of(pcs), weight, raevbrn, rlrs,
         raedyrs, r10sibs, edu, income_resid, income_med, married, 
         agefb, agefbn, rabyear, born, raehsamp, raestrat, 
         parent_weight, ragender, adhd_score, neurot, r10diabe, 
         r10hearte, r10cesd, age_smoked, anxiety, r10cogtot, 
         rlrs_div_h10gkid, child_weight)

@@

r_child <- svydesign(
    id = ~ raehsamp,
		strata = ~ raestrat,
		weights = ~ child_weight, 
		nest = TRUE,
		data = rand |> drop_na(child_weight) 
		)

@@

@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from @becker1976child. The theory has two components.

@@

<!--#  TE sources of pgs are not mentioned in the data section. I've added that since we are now using two sources -->

@@

For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository. For some traits, polygenic scores were created from European ancestry GWAS and GWAS with samples of different ancestry. We choose to use polygenic scores trained only on individuals of European ancestry. We discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. we also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs_white)` scores for the white participants and `r length(pgs_black)` for the black participants. PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score. In all regressions using PGS, we control for ten within-ethnicity principal components of the DNA array data.

The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by mean number of children of people born in the same year. RLRS is calculated pooling ethnicities, i.e. treating them as members of the same biological population. The mean number of children of people born in the same year was calculated using sampling weights.

@@

# cache seems to cause trouble even with cache-lazy: FALSE

@@

    form_pgs_gkids <- map(pgs, 
                       \(x) reformulate(c(x, pcs), response = "rlrs_div_h10gkid"))
  mod_pgs_gkids[[eth]] <- map(form_pgs_gkids,
                                    \(x) svyglm_quiet(x, r_child_eth))

@@

# TE bonferroni correction might make more sense focusing on white tests
# Either way seems to produce the same result

n_tests_white <- nrow(tidy_pgs$white)
n_sig_bonf_white <- sum(tidy_pgs$white$p.value < 0.05/n_tests)

n_tests_black <- nrow(tidy_pgs$black)
n_sig_bonf_black <- sum(tidy_pgs$black$p.value < 0.05/n_tests)

@@

```{r}
#| label: stat-bootstrap-eth-cor
#| cache: true
#| cache-lazy: false
#| eval: false

@@

<!--# TE edited because 3 scores are bonferroni significant in whites. As of writing the same same three scores were signficiant regardlesss of whether we pool the blacks and whites in the bonferroni calculation.-->

We estimated coefficients of PGS on RLRS among black and white respondents separately. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample. Standard errors are large because of the relatively low sample sizes. `r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests`. The scores are age at first birth, educational attainment and attention deficit disorder. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

```{r}
#| label: stat-bootstraps
#| cache: true
#| cache-lazy: false
#| cache-comments: false

@@

  cors_fert_edu[[eth]] <- 
    withReplicates(rand_boot, theta = calc_cors_edu, var = "rlrs")
  cors_sibs_edu[[eth]] <- 
    withReplicates(r_parents_boot, theta = calc_cors_edu, var = "r10sibs")
  cors_gkids_edu[[eth]] <- 
    withReplicates(r_child_boot, theta = calc_cors_edu, var = "rlrs_div_h10gkid")
}

@@

@fig-rlrs-edu plots each PGS's partial correlation with RLRS against its partial correlation with educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrap 95% C.I. `r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I. `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`). Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents.

@@

cor_no_outliers <- cor(tmp$estimate.sibs, tmp$estimate.edyrs)
# if (! cor_no_outliers < 0) stop("Correlation without outliers IS NOT neg.")
# TE correlation without outliers is not NEG
```

```{r}
#| label: stat-bootstrap-gen-cor
#| cache: true

@@

```{r}
#| label: stat-bootstrap-gen-cor-gkids
#| cache: true

@@

f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
f_gkids <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs_div_h10gkid"))

estimate_gen_cor <- function (w, data) {
  data$w <- w # avoid w getting lost by map and lm
  m_resp <- map(f_resp, \(x) lm(x, data, weights = w))
  m_gkids <- map(f_gkids, \(x) lm(x, data, weights = w*raevbrn))

  coef_resp <- map2_dbl(m_resp, pgs, \(x, y) coef(x)[[y]])
  coef_gkids <- map2_dbl(m_gkids, pgs, \(x, y) coef(x)[[y]])
  c(gen_cor = cor(coef_resp, coef_gkids))
}

@@

<!--# TE Added grandkids and fixed minor typos -->

We can also examine natural selection in the preceding and succeeding generations, by using reported number of siblings and grandchildren respectively. We regress PGS on respondents' number of living siblings in 2010. Data for dead siblings has too many missing values to use. We reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations. Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education. Correlations are only significant and negative in the white respondents, although ethnicities have large standard errors (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`; blacks: correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`). There is a positive and significant correlation across generations, i.e. between PGS coefficients on number of siblings and number of children (among whites: `r gen_cor`,cbootstrapped 95% C.I. `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).

To examine selection in the succeeding generation, we divide the respondents' number of grandchildren by their number of children. In other words, we calculate the average number of children the respondent's children have. This measures the reproductive success in the second generation which we can call RS2. `r gkids_unusable_white` white respondents and `r gkids_unusable_black` black respondents report having grandchildren despite reporting having had no children. These individuals were removed for this test. In regression we reweight respondents by the number of children they have, since more fecund grandparents account for a larger proportion of the next generation. Older grandparents will have more time for their number of grandkids to accumulate. To deal with this time trend, we divide RS2 by the mean RS2 of respondents born in the same year, creating relative RS2 (RRS2).

We regress the respondent's (the grandparent's) PGS on RRS2. The resulting coefficient is only a proxy for selection in the next generation, not a perfect measure. The grandparent's PGS indicates, but does not determine the PGS of the parent. To know the expected parent's PGS, we would need both grandparent's PGS. Given that some respondents do not have a partner in the HRS and some have had children with multiple partners, such an approach is untenable in our sample. We expect the effect of the grandparent's PGS to also be dependent upon the level of assortative mating for the trait. For traits with high assortative mating, the grandparent's PGS will more strongly predict the parent's PGS, leading to a greater regression slope.

Appendix @fig-gkids plots coefficients on RRS2 versus coefficients on years of education. Correlations are significant for whites but not blacks. Standard errors were moderate in whites, but large in blacks (whites: correlation `r cors_gkids_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_gkids$white[1]` to `r ci_boot_gkids$white[2]`; blacks: correlation `r cors_gkids_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_gkids$black[1]` to `r ci_boot_gkids$black[2]`). There is a strong positive correlation between the parent and grandchild regressions, i.e. between PGS coefficients on RRS2 and RLRS (among whites: `r gen_cor_gkids`, bootstrapped 95% C.I. `r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`).

@@

```{r}
#| label: stat-x-bootstrap-cors
#| cache: true
#| cache-lazy: false
#| eval: false

@@

```{r}
#| label: stat-x-bootstrap-effect-size
#| cache: true
#| cache-lazy: false
#| cache-comments: false

n_reps_x_es <-  199

@@

```{r}
#| label: stat-bootstrap-agefb
#| cache: true

n_reps_agefb <-  199

@@

stopifnot(insignificant(confint(cor_agefb)[1,]))
ci_discordant <- confint(cor_agefb)[2,]
# significantly different from 1/2?
stopifnot(ci_discordant[1] <= length(pgs)/2 && 
            ci_discordant[2] >= length(pgs)/2)

@@

# selection effects controlling for age at first birth
# 
form_pgs_control_agefb <- map(pgs, \(x) reformulate(c(x, "agefbn", pcs), 
                                                    response = "rlrs"))
mod_pgs_control_agefb <- map(form_pgs_control_agefb,
                             \(x) svyglm_quiet(x, rand, subset = 
                                                 ethnicity == "white"))
tidy_pgs_control_agefb <- map(mod_pgs_control_agefb, broom::tidy) |> 
                          list_rbind() |> 
                          filter(term %in% pgs)

@@

```{r}
#| label: stat-x-bootstrap-birthyear
#| cache: true

n_reps_birthyear <-  199

@@

@@

where \$h\^2\$ is the heritability of the PGS target phenotype and \$R\^2\_{PGS}\$ is the \$R\^2\$ of the measured PGS on the target phenotype [@becker2021resource]. Moving from error correction in a univariate model to a multivariate model requires a slightly more complex correction, although it uses the same parameters. To do this we use the error-corrected estimator developed by \@becker2021resource.

SNP- or chip-heritabilities and $R^2_{PGS}$ are calculated by @becker2021resource for scores in the Polygenic Index Repository. The authors use GCTA to estimate heritability. When the corresponding phenotype is not available in the HRS to estimate $R^2_{PGS}$, we use parameters estimated by the authors in the Wisconsin Longitudinal Study instead. We also perform error correction with twin heritabilites. To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^2]. We remove polygenic scores with $R^2_{PGS} > 0.005$ to focus on PGS with adequate power. This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction.

@@

@fig-true-effects reports the error corrected estimates of selection. As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean. Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero. Age of First Birth has a twin corrected heritability of 0.15, implying that it maybe under fast selection. The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true \$h\^2\$ , the true \$R\^2\_{PGS}\$, limitations of the within-sample phenotypes, noise from correlated environments, and for twin-heritability, different relationships with fertility among variants not measured on the chip. Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.

Results from a few traits show substantial uncertainty regarding the true effect size. For asthma, when chip heritability is used, the effect is positive and significant, but negative and significant when twin heritability is used! Such a large disparity is because @becker2021resource estimate the chip heritability at 0.015 and we use a twin heritability 0f 0.55. For personality traits, chip heritability is often low (e.g. $<0.05$) but twin heritabilities are often much higher ($≈0.4$) implying substantial missing heritability. For these traits, the confidence intervals on twin heritabilities are enormous. Error-corrected effect sizes and the $R^2$ and $h^2$ parameters used can be found on the Github page for this paper.

<!--# TE The lay public want to know how much IQ is changing. So I've added in a calculation of this. -->

>>>>

---
title: "Natural Selection Across Three Generations of Americans"
author: 
  - David Hugh-Jones
  - Tobias Edwards
abstract: |
  We investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  partially support the economic theory of fertility as an explanation for 
  natural selection: among both black and white respondents,
  scores which correlate negatively (positively) with education are
  selected for (against). Selection coefficients are larger among low-income
  and unmarried parents, but not among younger parents or those with less 
  education. Using respondents' number of siblings and grandchildren we 
  were able to examine selection across three generations, finding the 
  results consistent over time. We also estimate effect sizes 
  corrected for noise in the polygenic scores, finding selection
  for some health traits to be similar in magnitude to cognitive traits. 
date: "January 2024"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

```{r}
#| label: setup
#| cache-lazy: false

# TE One time I tried to render the document with cache on I got the error message on the first chunk: "Error in `lazyLoadDBinsertVariable()`: ! long vectors not supported yet: connections.c:6028" Added cache-lazy: false which seems to fix things

options(digits = 3)
set.seed(271075)

@@

pgirep <- readr::read_tsv("data/pgi-repository/HRS_PGIrepo_v1.0.txt",
                          show_col_types = FALSE)
names(pgirep)[1:2] <- c("hhid","pn")

@@

# rand family respondents file
# Done just once:
# rand_fam <- haven::read_dta(
#   "data/randhrsfam1992_2018v2_STATA/randhrsfamr1992_2018v2.dta")
# saveRDS(rand_fam, "data/randhrsfam1992_2018v2_STATA/rand_fam.rds")
rand_fam <- readRDS("data/randhrsfam1992_2018v2_STATA/rand_fam.rds")

# only selected columns to avoid creating suffixes when we join with
# rand.
# Age of oldest kid 2016; age of oldest kid 2010; grandchildren; num siblings who died
rand_fam <- rand_fam |> select(hhid, pn, h13ageokid, h10ageokid, h13gkid, r10sbdied)

@@

# Replacing 0 weights with NA because it can create odd problems
rand$weight <- na_if(rand$weight, 0)

@@

# TE this is average number of kids that your kids have
rand$div_h13gkid <- rand$h13gkid / rand$raevbrn

# Some people report having grandkids, but having had no children. Weird!
# This code subsets the data by ethnicity then counts all observations
# where grandkids is greater than 0 but number of kids is 0
gkids_unusable_white <- rand |> 
  filter(ethnicity == "white", h13gkid > 0, raevbrn == 0) |> 
  nrow()
gkids_unusable_black <- rand |> 
  filter(ethnicity == "black", h13gkid > 0, raevbrn == 0) |> 
  nrow()

# if grandkids but no kids div_gkid is infinite. Replacing with NA
rand$div_h13gkid[is.infinite(rand$div_h13gkid)] <- NA

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_weight = mean(weight, na.rm = TRUE),
    mean_child_weight = mean(child_weight, na.rm = TRUE),
    mean_raevbrn = mean(raevbrn*weight/mean_weight, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn,
    mean_h13gkid = mean(h13gkid*weight/mean_weight, na.rm = TRUE),
    rlrs_h13gkid = h13gkid/mean_h13gkid,
    mean_div_h13gkid = mean(div_h13gkid*child_weight/mean_child_weight, 
                            na.rm = TRUE), 
    rlrs_div_h13gkid = div_h13gkid/mean_div_h13gkid
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    r10hearte = as.numeric(r10hearte),
    r10diabe = as.numeric(r10diabe),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    born = chop(rabyear, 1942, labels = c("pre-1942", "post-1942")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )

@@

rand <- rand |> 
  select(ethnicity, all_of(pgs), all_of(old_pgs), 
         all_of(pcs), weight, raevbrn, rlrs,
         raedyrs, r10sibs, edu, income_resid, income_med, married, 
         agefb, agefbn, rabyear, born, raehsamp, raestrat, 
         parent_weight, ragender, adhd_score, neurot, r10diabe, 
         r10hearte, r10cesd, age_smoked, anxiety, r10cogtot, 
         rlrs_div_h13gkid, child_weight, h13ageokid)

@@

r_child <- svydesign(
    id = ~ raehsamp,
		strata = ~ raestrat,
		weights = ~ child_weight, 
		nest = TRUE,
		data = rand |> drop_na(child_weight) |>
		        filter(!is.na( rlrs_div_h13gkid), h13ageokid > 39)
		)

@@

@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from @becker1976child. The theory has two components:

@@

@@

For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository. For some traits, polygenic scores were created from both European ancestry and multiple ancestry GWAS. We choose to use polygenic scores trained only on individuals of European ancestry. We discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. we also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs_white)` scores for the white participants and `r length(pgs_black)` for the black participants. PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score. In all regressions using PGS, we control for ten within-ethnicity principal components of the DNA array data.

The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by the mean number of children of people born in the same year. RLRS is calculated pooling ethnicities, i.e. treating them as members of the same biological population. The mean number of children of people born in the same year was calculated using sampling weights.

@@

# cache seems to cause trouble even with cache-lazy: FALSE 
options("survey.lonely.psu" = "average")

@@

    form_pgs_gkids <- map(pgs, 
                       \(x) reformulate(c(x, pcs), response = "rlrs_div_h13gkid"))
  mod_pgs_gkids[[eth]] <- map(form_pgs_gkids,
                                    \(x) svyglm_quiet(x, r_child_eth))

@@

n_tests_white <- nrow(tidy_pgs$white)
n_sig_bonf_white <- sum(tidy_pgs$white$p.value < 0.05/n_tests_white)

n_tests_black <- nrow(tidy_pgs$black)
n_sig_bonf_black <- sum(tidy_pgs$black$p.value < 0.05/n_tests_black)

@@

```{r}
#| label: stat-bootstrap-eth-cor
#| cache: false
#| cache-lazy: false
#| eval: false

@@

We estimated coefficients of PGS on RLRS among black and white respondents separately. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample. Standard errors are large because of the relatively low sample sizes. In the white sample, `r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`. The scores are age at first birth, educational attainment, ADHD, and self-rated health. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

```{r}
#| label: stat-bootstraps
#| cache: false
#| cache-lazy: false
#| cache-comments: false

@@

  cors_fert_edu[[eth]] <- 
    withReplicates(rand_boot, theta = calc_cors_edu, var = "rlrs")
  cors_sibs_edu[[eth]] <- 
    withReplicates(r_parents_boot, theta = calc_cors_edu, var = "r10sibs")
  cors_gkids_edu[[eth]] <- 
    withReplicates(r_child_boot, theta = calc_cors_edu, var = "rlrs_div_h13gkid")
}

@@

@fig-rlrs-edu plots each PGS's regression beta on RLRS against its regression beta on educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrap 95% C.I. `r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I. `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`). Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents.

@@

cor_no_outliers <- cor(tmp$estimate.sibs, tmp$estimate.edyrs)
# if (! cor_no_outliers < 0) stop("Correlation without outliers IS NOT neg.")
# TE correlation without outliers is not NEG
# But it is not negative anyway now
```

```{r}
#| label: stat-bootstrap-gen-cor
#| cache: false

@@

```{r}
#| label: stat-bootstrap-gen-cor-gkids
#| cache: false

@@

f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
f_gkids <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs_div_h13gkid"))

estimate_gen_cor <- function (w, data) {
  data$w <- w # avoid w getting lost by map and lm
  m_resp <- map(f_resp, \(x) lm(x, data, weights = w))
  m_gkids <- map(f_gkids, \(x) lm(x, data |> filter(h13ageokid > 39), weights = w*raevbrn))
  # data subset so respondent's kids have 1+ married kids
  coef_resp <- map2_dbl(m_resp, pgs, \(x, y) coef(x)[[y]])
  coef_gkids <- map2_dbl(m_gkids, pgs, \(x, y) coef(x)[[y]])
  c(gen_cor = cor(coef_resp, coef_gkids))
}

@@

We can also examine natural selection in the preceding and succeeding generations, by using reported number of siblings and grandchildren respectively. We regress PGS on respondents' number of living siblings in 2010. Data for dead siblings has too many missing values to use. We reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations. Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education. Correlations are only significant and negative in the white respondents, although ethnicities have large standard errors (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`; blacks: correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`). There is a positive and significant correlation across generations, i.e. between PGS coefficients on number of siblings and number of children (among whites: `r gen_cor`, bootstrapped 95% C.I. `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).

To examine selection in the succeeding generation, we divide the respondents' number of grandchildren by their number of children. In other words, we calculate the average number of children the respondent's children have. This indicates reproductive success in the second generation which we can call RLRS2. `r gkids_unusable_white` white respondents and `r gkids_unusable_black` black respondents report having grandchildren despite reporting having reported no children. These individuals were removed for this test. In regression we reweight respondents by the number of children they have, since more fecund grandparents account for a larger proportion of the next generation.

Older grandparents will have more time for their number of grandkids to accumulate. To better deal with this time trend, we subset the data to parents whose oldest child was $40$ years old or older in $2016$, when the number of grandchildren was recorded. Appendix @fig-offspring-cohorts shows the distribution of the birth year of the oldest child, adjusted for sampling weights. A majority of the observations are after $1960$, the most recent year the respondents were born in. This reassures us that we are observing reproduction in later generations.

We regress the respondent's (the grandparent's) PGS on RLRS2. The resulting coefficient is only a proxy for selection in the next generation, not a perfect measure. The grandparent's PGS indicates, but does not determine the PGS of the parent. To know the expected parent's PGS, we would need both grandparent's PGS. Given that some respondents do not have a partner in the HRS and some have had children with multiple partners, such an approach is untenable in our sample. We expect the effect of the grandparent's PGS to also be dependent upon the level of assortative mating for the trait. For traits with high assortative mating, the grandparent's PGS will more strongly predict the parent's PGS, leading to a greater regression slope.

Appendix @fig-gkids plots coefficients on RLRS2 versus coefficients on years of education. Correlations are significant for whites but not blacks. Standard errors were large in both ethnicites (whites: correlation `r cors_gkids_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_gkids$white[1]` to `r ci_boot_gkids$white[2]`; blacks: correlation `r cors_gkids_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_gkids$black[1]` to `r ci_boot_gkids$black[2]`). There is a positive correlation between PGS coefficients on RLRS2 and RLRS (among whites: `r gen_cor_gkids`, bootstrapped 95% C.I. `r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`), indicating some stability in selection over time.

@@

```{r}
#| label: stat-x-bootstrap-cors
#| cache: false
#| cache-lazy: false
#| eval: false

@@

```{r}
#| label: stat-x-bootstrap-effect-size
#| cache: false
#| cache-lazy: false
#| cache-comments: false

n_reps_x_es <-  199
pgs <- pgs_white

@@

```{r}
#| label: stat-bootstrap-agefb
#| cache: false

n_reps_agefb <-  199
pgs <- pgs_white

@@

stopifnot(insignificant(confint(cor_agefb)[1,]))
ci_discordant <- confint(cor_agefb)[2,]
# significantly different from 1/2?
stopifnot(ci_discordant[1] <= length(pgs)/2 && 
            ci_discordant[2] >= length(pgs)/2)
# TE THEY ARE DISCORDANT

@@

# selection effects controlling for age at first birth
pgs <- pgs_white

form_pgs_control_agefb <- map(pgs, \(x) reformulate(c(x, "agefbn", pcs), 
                                                    response = "rlrs"))
mod_pgs_control_agefb <- map(form_pgs_control_agefb,
                             \(x) svyglm_quiet(x, rand, subset = 
                                                 ethnicity == "white"))
tidy_pgs_control_agefb <- map(mod_pgs_control_agefb, broom::tidy) |> 
                          list_rbind() |> 
                          filter(term %in% pgs)

@@

```{r}
#| label: stat-x-bootstrap-birthyear
#| cache: false

n_reps_birthyear <-  199
pgs <- pgs_white

@@

```{r}
#| label: ADHD
#| cache: false

# Testing the R2 of the ADHD PGS
rand$variables |>
  with(cor.test(adhd_score, ADHD, use = "pairwise.complete")) -> adhd_cor

@@

where $h^2$ is the heritability of the PGS target phenotype and $R^2_{PGS}$ is the coefficient of determination of the measured PGS on the target phenotype [@becker2021resource]. Moving from error correction in a univariate model to a multivariate model, controlling for principal components of genetic array data, requires a slightly more complex correction, although it uses the same parameters. To do this we use the error-corrected estimator developed by @becker2021resource.

SNP- or chip-heritabilities and $R^2_{PGS}$ are calculated by @becker2021resource for scores in the Polygenic Index Repository. The authors use GCTA to estimate heritability. When the corresponding phenotype is not available in the HRS to estimate $R^2_{PGS}$, we use parameters estimated by the authors in the Wisconsin Longitudinal Study instead. We also perform error correction with twin heritabilites. To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^2]. We remove polygenic scores with $R^2_{PGS} < 0.005$ to focus on PGS with adequate power. This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction[^3].

@@

[^3]: @becker2021resource estimate $R^2_{PGS}$ for ADHD using binary, yes or no, items of ADHD symptoms. These were PV001-PV018 in the 2016 core survey. They estimated an $R^2_{PGS}$ of $0.003$. When we attempt to replicate it we find an $r_{PGS}$ of `r adhd_cor$estimate`, implying an $R^2_{PGS}$ of `r adhd_cor$estimate^2`. Although this would pass our requirement of $R^2_{PGS} > 0.005$, the correlation was not significant (p = `r adhd_cor$p.value`) and the confidence intervals were large 95% C.I. `r adhd_cor$conf.int[1]` to `r adhd_cor$conf.int[2]` owing to the fact that few respondents were given the ADHD items N = `r adhd_cor$parameter + 2`. Given the great uncertainty over the true correlation we are unable to justify using the error-corrected estimator with the ADHD polygenic score.
